Visualizing probability distributions across bivariate nonlinear temporal granularities

### 2020-02-24 main.Rmd
content cut - Section 3 - iteration 1

   Visualizing data across granularities which are either single-order-up or multiple-order-up or periodic/aperiodic helps us to understand periodicities, pattern and anomalies in the data. Because of the large volume of data available, using displays of probability distributions conditional on one or more granularities is a potentially useful approach. This work provides tools for creating granularities and exploring the associated time series within a tidy workflow, so that probability distributions can be examined using the range of graphics available in ggplot2 [@Wickham2009pk].
bibliography: [bibliography.bib]
preamble: >
  \usepackage{mathtools,amssymb,booktabs,longtable,todonotes}
  \def\mod{~\text{mod}~}
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    fig_height: 5
    fig_width: 8
    fig_caption: yes
    dev: "pdf"
    keep_tex: yes
---

@@

We propose a formalism of nonlinear time granularities through the tsibble [@wang2019tsibble] framework of organizing temporal data. A tsibble consists of an index, key and measured variables. An index is a variable with inherent ordering from past to present and a key is a set of variables that define observational units over time. In a tsibble, each observation (row) should be uniquely identified by index and key.

@@

Nonlinear time granularities with periodicities can be constructed by relating two linear granularities. The mappings between these linear granularities can be regular or irregular. For example, a regular mapping exists between minutes and hours (where 60 minutes always add up to 1 hour), while an irregular mapping exists between days and months (as the number of days per month ranges from 28 to 31). Mappings will be regular if linear granularities are formed by a grouping operation, but granularities formed by the altering-tick operation will be irregular. Nonlinear time granularities are referred to as **circular** if mappings between linear time granularities are regular and **quasi-circular** if the mapping is irregular. Examples of common circular granularities include hour of the day, and day of the week, whereas examples for quasi-circular granularities can be day of the month or week of the month.

@@

\autoref{fig:circular-dow} depicts the circular granularity "day of the week", constructed from two linear granularities "day" (the bottom granularity) and "week". The granules that overlap share elements from the underlying time domain. In this example, $P=7$, and the circular granularity $C_{\text{day}, \text{week}}$ representing day-of-week consists of the repeated pattern $\{L(0), L(1), L(2),$ $L(3), L(4), L(5), L(6)\}$.

@@

In general, any circular granularity relating two linear granularity can be expressed as $C_{(G, H)}(z) = L(\lfloor z/P(B,G) \rfloor\mod P(G,H))$, where linear granularity $H$ is periodic with respect to linear granularity $G$ with regular mapping such that the number of granules of $G$ in each granule of $H$ is $P(G,H)$. Table \ref{tab:definitions} shows representation of circular granularities $C_i$ relating two linear granularities with $P_i$ being the number of granules of the finer granularity in each granule of the coarser granularity and $L_i$ is the associated label mapping. It is possible that none of the two linear granularities are bottom granularities. But the representation of the resultant circular granularity will be a function of the index set. The bottom granularity is assumed to be minutes.

@@

\begin{equation} \label{eq:eq4}
\begin{split}
Q_{B,G'}(8) & = L(8 - \sum_{w=0}^{3-1}\vert T_{w \mod 2}\vert) ,\quad since \quad 8 \in T_{3} \\
  & = L(8 - \sum_{w=0}^{2}\vert T_{w\mod 2}\vert) \\
  & = L(8 - \vert T_{0 \mod 2}\vert - \vert T_{1 \mod 2}\vert - \vert T_{2 \mod 2}\vert) \\
  & = L(8 - 2*\vert T_{0}\vert - \vert T_{1}\vert) \\
  & = L(8 - 2*2 - 3) \\
  & = L(1) \\
\end{split}
\end{equation}

\begin{equation} \label{eq:eq5}
\begin{split}
Q_{B,H'}(10) & = L(10 - \sum_{w=0}^{1-1}\vert T_{w \mod 2}\vert) ,\quad since \quad  10 \in T_{1}  \\
  & = L(10 - \sum_{w=0}^{0}\vert T_{w\mod 2}\vert) \\
  & = L(10 - \vert T_{0 \mod 2}\vert) \\
  & = L(10 - \vert T_{0}\vert) \\
  & = L(10 - 7) \\
  & = L(3) \\
\end{split}
\end{equation}

**Discussion**: If linear granularity $G'$ is periodical with respect to $B$ with irregular mapping, then the following holds true:

- $B \trianglelefteq G'$
- there exist $R'$, $P' \in \mathbb{Z}_+$ such that if $G'(w) = \bigcup_{z \in T_w}B(z)$ then $$G'(w) = \bigcup_{z \in T_w \mod R'}B(P'*\lfloor w/R' \rfloor + z)$$

from definition \ref{def:periodical}

and,  number of granules of $B$ in each granule of $G'$ is not a constant (irregular mapping)

@@

It can be noted here that if a linear granularity $G'$ is quasi-periodic (Definition \autoref{def:quasi}) with respect to $B$, then \autoref{eq:quasi} can be modified as follows to account for exceptions $E = [e_{begin}, e_{end}]$ (by Definition \autoref{def:quasi}).

@@

for $z \in T_k$ and $e_k$ is the number of exceptions in $\bigcup\limits_{w = 0}^{k} T_w$ and $E_{u} = [e_{begin}(u), e_{end}(u)]$.

## Aperiodic {#sec:aperiodic-gran-def}

>>>>

   Visualizing data across granularities which are either single-order-up or multiple-order-up or periodic/aperiodic helps us to understand periodicities, pattern and anomalies in the data. Because of the large volume of data available, using displays of probability distributions conditional on one or more granularities is a potentially useful approach. This work provides tools for creating granularities and exploring the associated time series within a tidy workflow, so that probability distributions can be examined using the range of graphics available in ggplot2 [@Wickham2009pk].
bibliography: [bibliography.bib]
preamble: >
  \setlength {\marginparwidth }{2cm}
  \usepackage{mathtools,amssymb,booktabs,longtable,todonotes}
  \def\mod{~\text{mod}~}
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    fig_height: 5
    fig_width: 8
    fig_caption: yes
    dev: "pdf"
    keep_tex: yes
---

@@

We propose a formalism of nonlinear time granularities through the tsibble [@wang2019tsibble] framework of organizing temporal data. A tsibble consists of an index, key and measured variables. An index is a variable with inherent ordering from past to present and a key is a set of variables that define observational units over time. In a tsibble, each observation (row) is uniquely identified by index and key.

@@

Nonlinear time granularities with periodicities can be constructed by relating two linear granularities. The mappings between these linear granularities can be regular (minutes and hours) or irregular (days and months). <!--For example, a regular mapping exists between  (where 60 minutes always add up to 1 hour), while an irregular mapping exists between  (as the number of days per month ranges from 28 to 31).--> A grouping operation in Calendar algebra could lead to a regular mapping, whereas a altering-tick operation could result in an irregular mapping. Nonlinear time granularities are referred to as **circular** if mappings are regular and **quasi-circular** if the mapping is irregular. Examples of common circular granularities include hour of the day, and day of the week, whereas examples for quasi-circular granularities can be day of the month or week of the month.

@@

\autoref{fig:circular-dow} depicts the circular granularity "Day-of-Week", constructed from linear granularities "day" (the bottom granularity) and "week". The granules that overlap share elements from the underlying time domain. In this example, $P=7$, and the circular granularity $C_{\text{day}, \text{week}}$ representing day-of-week consists of the repeated pattern $\{L(0), L(1), L(2),$ $L(3), L(4), L(5), L(6)\}$.

@@

In general, any circular granularity relating two linear granularity can be expressed as $C_{(G, H)}(z) = L(\lfloor z/P(B,G) \rfloor\mod P(G,H))$, where linear granularity $H$ is periodic with respect to linear granularity $G$ with regular mapping such that the number of granules of $G$ in each granule of $H$ is $P(G,H)$. Table \ref{tab:definitions} shows representation of circular granularities $C_i$ relating two linear granularities with $P_i$ being the number of granules of the finer granularity in each granule of the coarser granularity and $L_i$ is the associated label mapping. <!--The representation of the resultant circular granularity will be a function of the index set.--> The bottom granularity is assumed to be minutes in this case.

@@

\begin{equation} \label{eq:eq4}
\begin{split}
Q_{B,G'}(8) & = L(8 - \sum_{w=0}^{3-1}\vert T_{w \mod 2}\vert) ,\quad since \quad 8 \in T_{3} \\
  & = L(1) \\
\end{split}
\end{equation}

<!-- CONTENT CUT -->
  <!-- & = L(8 - \sum_{w=0}^{2}\vert T_{w\mod 2}\vert) \\ -->
  <!-- & = L(8 - \vert T_{0 \mod 2}\vert - \vert T_{1 \mod 2}\vert - \vert T_{2 \mod 2}\vert) \\ -->
  <!-- & = L(8 - 2*\vert T_{0}\vert - \vert T_{1}\vert) \\ -->
  <!-- & = L(8 - 2*2 - 3) \\ -->

  <!-- & = L(10 - \sum_{w=0}^{0}\vert T_{w\mod 2}\vert) \\ -->
  <!-- & = L(10 - \vert T_{0 \mod 2}\vert) \\ -->
  <!-- & = L(10 - \vert T_{0}\vert) \\ -->
  <!-- & = L(10 - 7) \\ -->

\begin{equation} \label{eq:eq5}
\begin{split}
Q_{B,H'}(10) & = L(10 - \sum_{w=0}^{1-1}\vert T_{w \mod 2}\vert) ,\quad since \quad  10 \in T_{1}  \\
  & = L(3) \\
\end{split}
\end{equation}

**Discussion**: If linear granularity $G'$ is periodical with respect to $B$ with irregular mapping, then there exist $R'$, $P' \in \mathbb{Z}_+$ such that if $G'(w) = \bigcup_{z \in T_w}B(z)$ then $$G'(w) = \bigcup_{z \in T_w \mod R'}B(P'*\lfloor w/R' \rfloor + z)$$ (from definition \ref{def:periodical})

<!-- - $B \trianglelefteq G'$ -->

-

<!-- and,  number of granules of $B$ in each granule of $G'$ is not a constant (irregular mapping) -->

@@

It can be noted here that if a linear granularity $G'$ is quasi-periodic (Bettini2000-vy) <!--(Definition \autoref{def:quasi})--> with respect to $B$, then \autoref{eq:quasi} can be modified as follows to account for exceptions $E = [e_{begin}, e_{end}]$.<!--(by Definition \autoref{def:quasi}).-->

@@

for $z \in T_k$ and $e_k$ is the number of exceptions in $\bigcup\limits_{w = 0}^{k} T_w$ and $E_{u} = [e_{begin}(u), e_{end}(u)]$.
## Aperiodic {#sec:aperiodic-gran-def}
